---
title: 'esvis: Part 1'
author: Daniel Anderson
date: '2017-11-05'
slug: esvis-part-1
categories: []
tags:
  - data_vis
output: html_document
---

This is the first of a series of posts to introduce my new *esvis* R package,
why I think it's important, and some of its capabilities. As of this writing
the current version on CRAN is 0.0.1, so it's obviously still fresh and may
have some bugs. If you find any, please let me know. You can install the package
like you would any other on R

```{r install_esvis_cran, eval = FALSE}
install.packages("esvis")
```

or if you'd prefer the sometimes buggy but more feature-heavy development 
version, install from github with devtools. 

```{r install_esvis_github, eval = FALSE}
# install.packages("devtools") # only if not previously installed
devtools::install_github("DJAnderson07/esvis")
```

## Overview
The overall purpose of the package is to visualize distributional differences.
Often, we think about distributional differences in terms of effect sizes, which
is certainly a good method, but these effect sizes are not without limitations.
They are *summary* indicators of the distributional differences, generally 
providing a standardized measure of the difference between the means. Cohen's 
*d* is perhaps the most common effect size metric, and is defined by the 
difference in the means between two distributions divided by the pooled standard 
deviation. 

In some cases, we may be interested in the difference between two 
distributions at other points in the scale, which may be more relevant to 
applied use. For example, students in public school systems across the United
States take statewide achievement tests in reading and mathematics. These scores 
are evaluated relative to performance
level classifications (e.g., "needs improvement", "proficient", "advanced"). In
these cases, it may be more policy-relevant to evaluate the differences between
the distributions (different groups of students) at these cut points, rather 
than at the means. However, all of this comes with a caveat - if the two (or 
more) distributions are both normally distributed with the same variance, the 
difference at the means will equal the difference at any point in the scale. In 
this case, we could just as well evaluate differences at the mean and we'd 
understand the  difference at the cut-point for "proficient" or any other point
on the scale. But, real data rarely work out this way and sometimes important
and meaningful differences can occur at different point in the scale. In this
case, any effect size measure will be at least somewhat insufficient, because
it is necessarily trying to summarize the entire difference between the 
distributions with a single number - but this difference may depend up on the
scale location.

All of the previous paragraph is really just a big run-up to hopefully convince
you that if you're interested in the difference between any two empirical 
distributions, it's likely helpful to *visualize* those differences, because you
may find that the difference is not consistent across the scale. This can occur
when the shape of the distributions differ, and/or when the variance between the
two distributions are more that moderately different.

## PP plots - and implementation in *esvis*
One way to visualize distributional differences is through 
probability-probability, or PP, plots. PP plots map the empirical cumulative
distribution function (CDF) from a reference distribution to the empirical CDF
of a focal distribution. These CDFs can be thought of in terms of percentile 
ranks - how different is the 10th percentile for students who are and are not
eligible for free or reduced price lunch? If there is no difference between the
distributions, the PP Curve will follow a diagonal line, usually displayed on 
the PP plot for reference. The extent to which the curve deviates from the 
reference line relates to the magnitude of the differences between the two 
distributions. Importantly, this allows for the investigation of the size of
the distributional differences across the full scale.

The *esvis* package will produce PP plots quickly using standard and consistent
syntax of the type `pp_plot(outcome ~ grouping_factor, data)`. One of the nice
features of the package is that it the grouping factor can have many levels and
the package will choose a reference distribution and plot multiple lines 
relative to that single reference distribution (which can be changed easily).
One of the datasets that ships with the development version (soon to be on CRAN)
is called `benchmarks` and includes seasonal (fall, winter, spring) assessment
data on students in Grades 3-5. These are synthetic data, simulated from 
empirical data. The properties of the synthetic data match the empirical data 
well. Below is a sample of the data

```{r data, echo = FALSE}
knitr::kable(head(esvis::benchmarks))
```

To produce a basic PP plot evaluating the differences in reading achievement
between students who are and are not eligible for free or reduced price lunch,
we would run

```{r pp_plot_frl}
library(esvis)
pp_plot(reading ~ frl, benchmarks)
```

Note that I'm using the development version, which looks slightly different
than the current CRAN release. The plot is annotated with some additional 
features that I'll explain more fully in a later blog post. But for now,
we can see that there is a sizable difference between these groups, and
that difference appears relatively consistent.

Let's evaluate another relation, this time looking at the different English 
language learner (ELL) classifications. This dataset includes three ELL 
designations: (a) active, (b) monitor, and (c) non-ell. Active refers to 
students currently receiving services, monitor refers to students who previously
received services, and non-ell refers to students who never received services. 
In the code below, I've added one additional argument to force the reference
group to be students who never received ELL services.

```{r pp_plot_ell}
pp_plot(reading ~ ell, benchmarks, ref_group = "Non-ELL")
```

Notice in this plot there is actually a reversal of the effect for monitor 
students. On the lower end of the scale, Monitor students are actually 
out-performing non-ELL students, but this effect reverses at the top of the 
scale. A summary measure would not provide this type of information, but it
may be incredibly valuable for theory development. For example, for this finding
we may theorize that students with very low achievement receive a benefit from
essentially any additional attention, even if that attention is not directly
related to academics.

That's it for now. Future posts will talk about other visualizations, as well
as the estimation of different kinds of effect size. I'll likely have at least
one post discussing some extensions to the basic plots produced above.