---
title: esvis
author: Daniel Anderson
date: '2017-11-02'
slug: esvis
categories: [data_vis]
tags: [data_vis]
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(fig.width = 4.5)
```


Here's some more stuff, this time with my *esvsis* package.

Here's a sampling of the data we'll be playing with.
```{r data}
library(esvis)
knitr::kable(head(benchmarks))
```

Here's a probability-probability plot for English learners in reading.
```{r esvis, fig.width = 5}
pp_plot(reading ~ ell, benchmarks)
```

There's customizable themes too. Here' the built in dark theme.

```{r esvis_dark}
pp_plot(reading ~ ell, benchmarks, theme = "dark")
```

We can also look at binned effect size plots for a similar view of the data, 
but this time with Cohen's d. Let's look at it by quintile, and we'll use the 
tan theme this time.

```{r binned_quintile}
binned_plot(reading ~ ell, 
            data = benchmarks, 
            qtiles = seq(0, 1, 0.2),
            theme = "tan")
```

Okay. That's good enough for now.