---
title: Alluvial Diagrams with ggforce
author: Daniel Anderson
date: '2017-11-29'
slug: alluvial-plots-with-ggforce
categories: []
tags:
  - data_vis
---
Today I wanted to quickly share my first real attempt at making an alluvial 
diagram. For those not familiar (and I wasn't previously) an 
[alluvial](https://en.wikipedia.org/wiki/Alluvial_diagram) diagram is a type of
flow plot that is essentially equivalent to a 
[sankey](https://developers.google.com/chart/interactive/docs/gallery/sankey)
diagram. The difference is that while sankey diagrams show flow for different
categorical variables, alluvial plots show change over time. To produce the 
alluvial diagram below, I'll be using the development version of the excellent
[ggforce](https://github.com/thomasp85/ggforce) package written by 
[Thomas Lin Pedersen](https://twitter.com/thomasp85), who's not only incredibly
talented, but also a good follow on twitter.

** insert plot **

## Context
In schools across the country students who are struggling academically may be
screened and evaluated to determine if they have a specific learning disability
(SLD). Unfortunately, the protocols for making these determinations are not
always consistent. We were interested, therefore, in examining the extent to 
which students shift into and out of the SLD classification over time. In this
particular example, we only have data for middle school students (Grades 6-8),
although 
[similar research](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.839.1628&rep=rep1&type=pdf)
has examined other grades. 

## The data
The actual dataset is rather large, and contains statewide testing records for
all students in the state of Oregon across Grades 6-8 for three cohorts. For
the purposes of this blogpost, I first simulated new data (so I could share it
here) using the 
[synthpop](https://cran.r-project.org/web/packages/synthpop/vignettes/synthpop.pdf) 
package, which is pretty impressive and perfect for this sort of purpose (I
unfortunately am not allowed to share the real data publicly). I then took a 
random sample of 5% of the students in the synthetic data so it wouldn't be so
large. The data are available [here](../data/synthetic_data2.csv), and you can 
read the file in directly using the following block of code, (which also cleans
up the names a bit and selects only the relevant variables).

```{r load_data, message = FALSE}
library(tidyverse)
d <- read_csv("/Users/Daniel/Teaching/guest_lecture_Gina/synthetic_data2.csv") %>% 
  janitor::clean_names() %>% 
  select(sid, grade, sld, ld33)
d
```
The *sld* column is a dummy vector indicating if the given student (*sid*) was
identified with a specific learning disability in that grade (1) or not (0). The
*ld33* column has an odd name, but basically states if the student was always,
never, or sometimes identified as SLD.

## The alluvial plot

We first need to install the development version of *ggforce*.
```{r install_ggforce, eval = FALSE}
devtools::install_github("thomasp85/ggforce")
```

Next, we need to do a bit of data prep. Below I:
* `filter()` for only students who were *ever* identified as SLD - i.e., I'm 
eliminating students who were always identified as Non-SLD, because that group 
encompasses the majority of students and makes the rest of the patterns difficult 
to see.

* Recode `ld` and `grade` so they are character vectors. This is not only 
helpful in making the plot look nice, but actually vital in getting it to 
produce (I failed over and over again because I skipped this step in my 
initial attempts).

* `spread` the data out so there's a column specifying the *ld* status for each
student in each grade.

* `count` the number of unique patterns across grades. This will be used to 
provide the widths of the alluvial plot.

* `mutate` to create a new pattern variable that states, in a single variable,
which "flow" students in that group were in.

Note - I saved this all in an object `pd`, which is a general convention I use
that stands for plot data.

```{r data_prep}
pd <- d %>% 
  filter(ld33 != "Never") %>%
  mutate(sld = ifelse(sld == 0, "Non-SLD", "SLD"),
         grade = paste0("Grade ", grade)) %>% 
  spread(grade, sld) %>% 
  group_by(`Grade 6`, `Grade 7`, `Grade 8`) %>% 
  count() %>% 
  mutate(Pattern = paste(`Grade 6`, `Grade 7`, `Grade 8`, sep = "/"))
pd
```

Now we're getting close, but the actual format `ggforce` needs the data in is a
bit more complex. Luckily, the package provides a helper function. We just need
to tell it, basically, which columns we want along the x-axis.

```{r gather_set_data}
library(ggforce)
pd <- pd %>% 
  gather_set_data(1:3)

pd
```

And now we're all ready to go. The set of functions *ggforce* provides to
produce the alluvial plot is `geom_parallel_sets`. The basic plot can now be 
produced with

```{r alluvial_plot1, fig.width = 9, fig.height = 7}
ggplot(pd, aes(x = x, id = id, split = y, value = n)) +
    geom_parallel_sets(aes(fill = Pattern), alpha = 0.6) +
    geom_parallel_sets_axes(axis.width = 0.1, fill = "gray70") +
    geom_parallel_sets_labels(color = 'White', size = 6)
```

And we're basically there!

A bit of cleaning up will make the plot even a bit nicer. Here I've used 
`theme_void` to remove most of the extras like axes, but added back in the
labels on the x-axis. I've also repositioned the legend, and it's debatable
whether it's really needed at all. Also, the 
`scale_x_discrete(expand = c(0.05,0.05))` argument is important and removes a 
bunch of white space on the left and right margins.

```{r alluvial_plot2, fig.width = 9, fig.height = 7}
ggplot(pd, aes(x = x, id = id, split = y, value = n)) +
  geom_parallel_sets(aes(fill = Pattern), alpha = 0.6) +
  geom_parallel_sets_axes(axis.width = 0.1, fill = "gray70") +
  geom_parallel_sets_labels(color = 'White', size = 6) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "", y = "") +
  scale_x_discrete(expand = c(0.05,0.05)) +
  theme_void() + 
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 0),
    legend.position = "bottom",
    legend.direction = "horizontal"
    )
```
